Concurrent Programming in Harmony

### 2021-12-25 book/paper.tex
Working on updating paper.tex to use new preprocessor

However, if the model checker does not find any bugs, we do not
know for sure that the algorithm works for lists of more than
five numbers or for lists that have values other than the numbers
1 through 5.  Still, we would expect that the likelihood that there
are bugs remaining in the sorting algorithm is small.
That said, it would be easy to write a program
that sorts all lists of up to five numbers correctly but fails to
do so for a list of 6 numbers.  (Hint: simply use an \textbf{if}
statement.)

@@

The $\textbf{assert}$ statement checks that the output is
correct.  If the program is correct, Harmony reports the size of the
``state graph'' (13 states in this case).  If not, Harmony also
reports what went wrong, typically by displaying a summary of an execution in
which something went wrong.

@@

Concurrent programming, aka multithreaded programming, involves multiple
threads
\index{thread}%
running in parallel while sharing variables.
\autoref{fig:progs} shows two programs.  Program (a) is sequential.
It sets <{shared}> to <{True}>, asserts that
<{shared}> = <{True}> and finally sets <{shared}> to <{False}>.
If you run the program through Harmony, it will not find any problems
because there is only one execution possible and 1) in that execution
the assertion does not fail and 2) the execution terminates.
Program (b) is concurrent---it executes methods <{f()}> and
<{g()}> in parallel.
If method <{g()}> runs and completes before <{f()}>, then
the assertion in <{f()}> will fail when <{f()}> runs.
This problem is an example of non-determinism: methods <{f()}>
and <{g()}> can run in either order.
In one order, the assertion fails, while in the other it does not.
But since Harmony will find all possible executions, it will find
the problematic one.

@@

\begin{problems}
\item \autoref{fig:incenter} shows an attempt at trying to fix the code of
    \autoref{fig:inc}.  Run it through Harmony and see what happens.  Based on
    the error output, describe in English what is wrong with the code by describing,
    in broad steps, how running the program can get into a bad state.
\item What if we moved line~5 of \autoref{fig:incenter} to after the \textbf{if}
    statement (between lines~7 and~8)?  Do you think that would work?  Run it through
    Harmony and describe either why it works or why it does not work.
\end{problems}

@@

\glsadd{sequential consistency}%
In 1981, Gary L.~Peterson came up with a beautiful solution to the mutual exclusion
problem, now known as ``Peterson's Algorithm''~\cite{Peterson81}.
The algorithm is an amalgam of the (incorrect) algorithms in
\autoref{fig:naiveflags} and \autoref{fig:naiveturn}, and is presented
in \autoref{fig:peterson}.
(The first line specifies that the \textit{flags} and <{turn}>
variables are assumed to satisfy \emph{sequential consistency}---it prevents
Harmony from complaining about data races involving these variables,
explained in \autoref{ch:spinlock}.)

@@

The HVM starts in an initial state in which there is only
one thread (<{__init__}>()) and its program counter is~0.
A \emph{trace}
\index{trace}%
is a sequence of steps starting from the initial state,
resulting in a sequence of states.
When making a step, there are two sources of non-determinism
\index{non-determinism}%
in Harmony.
% \footnote{Actually, there is a third type called
% an \emph{interrupt}, discussed in \autoref{ch:interrupts}.}
One is when
there is more than one thread that can make a step.  The other is
when a thread executes a <{choose}> operation and there is
more than one choice.
Because there is non-determinism, there are multiple possible traces.
We call a state \emph{reachable}
\index{reachable state}%
if it is either the initial state
or it can be reached from the initial state through a finite trace.
A state is final
when there are no threads left to make state changes.

@@

When running Harmony, this assertion will fail.
You can check the HTML output to see what happened.
Suppose thread 0 is at the critical section, $\mathit{flags}[0]$ = true,
$\mathit{turn} = 1$, and thread 1 just finished the step in line 7,
setting $\mathit{flags}[1]$ to true.  Then $C(0)$ is violated.
But it suggests a new property:
$\mathcal{G}(i) =
\mathtt{thread}(i)@\mathtt{cs} \Rightarrow \mathcal{C}(i) \lor \mathtt{thread}(1-i)@10$.
That is, if thread <{i}> is at the critical section, then
either $\mathcal{C}(i)$ holds or thread $1-i$ is about to execute line 10.
\autoref{fig:petersonproof} formalizes $\mathcal{G}(i)$ in Harmony.
The label \texttt{gate} refers to line 10, that is,
the step that sets <{turn}> to $1-i$.
You can run \autoref{fig:petersonproof} to determine
that $\mathcal{G}(i)$ is an invariant for $i = 0, 1$.
Moreover, if $\mathcal{F}(i)$ and $\mathcal{G}(i)$ both hold for $i = 0, 1$,
then mutual exclusion holds.  We can show this using proof by
contradiction.  Suppose mutual exclusion is violated and thus both threads are in
the critical section.  By $\mathcal{F}$ it must be the case that both
\textit{flags} are true.  By $\mathcal{G}$ and the fact that neither thread
is at label \texttt{gate}, we know that both $C(0)$ and $C(1)$ must hold.
This then implies that $\mathit{turn} = 0 \land \mathit{turn} = 1$, providing
the desired contradiction.

@@

First consider the case that thread~0 is at label <{cs}>.  If thread~0
were to take a step, then in the next state thread~0 would be no longer
at that label and $\mathcal{G}(0)$ would hold trivially over the next state.
Therefore we only need to consider a step by thread~1.
%
From $\mathcal{G}$ we know that one of the following three cases must hold before
thread~1 takes a step:
\begin{enumerate}
\item \textit{flags}[1] = <{False}>;
\item <{turn}> = 0;
\item thread~1 is at label \texttt{gate}.
\end{enumerate}

Let us consider each of these cases.
We have to show that if thread~1 takes a step, then one of those
cases must hold after the step.
In the first case, if thread~1 takes a step, there are two possibilities:
either $flags[1]$ will still be <{False}> (in which case the first case
continues to hold), or $flags[1]$ will be <{True}>
and thread~1 will be at label \texttt{gate} (in which case the third case
will hold).
We know that thread~1 never sets <{turn}> to 1, so
if the second case holds before the step, it will also hold after the step.
Finally, if thread~1 is at label \texttt{gate} before the step, then after
the step <{turn}> will equal 0, and therefore the second case will hold
after the step.

@@

A cool anecdote is the following.  When the author of Harmony had to teach
Peterson's Algorithm, he refreshed his memory by looking at the Wikipedia
page.  The page claimed that the following predicate is invariant:
if thread <{i}> is in the critical section, then $\mathcal{C}(i)$ (i.e.,
$\mathcal{G}$ without the disjunct that thread $1-i$ is at label \texttt{gate}).
We already saw that this is not an invariant.
(The author fixed the Wikipedia page with the help of Fred B.~Schneider.)

@@

Harmony is not an object-oriented language like Python is.  In Python,
you can pass a reference to an object to a method, and that method
can then update the object.  In Harmony, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this, as mentioned in \autoref{ch:harmonymachine},
each shared variable has an \emph{address}, itself a Harmony value.
\index{address}%
If <{x}> is a shared variable, then the expression <{?x}> is the address of <{x}>.
If a variable contains an address, we call that variable a \emph{pointer}.
\index{pointer}%
If <{p}> is a pointer to a shared variable, then the
expression \texttt{!}<{p}> is the value of the shared variable.
In particular, <{!?x = x}>.
This is similar to how C pointers work (\texttt{*\string&}<{x}> = <{x}>).

@@

The code in \autoref{fig:lockspec} also uses the Harmony <{when}>
statement.  A <{when}> statement waits until a time in which
condition holds (not necessarily the first time) and then executes the
statement block.  The ``\texttt{await} \textit{condition}'' statement is
the same as ``<{when}> \textit{condition}: <{pass}>''.
Combined with the <{atomically}> keyword, the entire statement
is executed atomically at a time that the condition holds.

@@

To support implementing locks,
many CPUs have an atomic ``test-and-set'' (TAS)
\index{test-and-set}%
\index{TAS}%
operation.
Method \texttt{test\_and\_set} in \autoref{fig:tas} shows its specification.
Here <{s}> points to a shared boolean variable and <{p}>
to a private boolean variable, belonging to some thread.
The operation copies the value of the shared variable to the
private variable (the ``test'')
and then sets the shared variable to <{True}> (``set'').

\autoref{fig:tas} goes on to implement mutual exclusion for
a set of <{N}> threads.
The approach is called \emph{spinlock},
\index{spinlock}%
because each thread is ``spinning'' (executing a tight loop) until
it can acquire the lock.
The program uses $\mathtt{N}+1$ boolean variables.
Variable <{shared}> is initialized to <{False}> while
\textit{private}[<{i}>] for each thread <{i}> is initialized to <{True}>.

An important invariant, $\mathcal{I}_1$, of the program is that at any time at most
one of these variables is <{False}>.
Another invariant, $\mathcal{I}_2(i)$, is that if thread <{i}> is
in the critical section,
then \textit{private}[<{i}>] = <{False}>.
Between the two (i.e., $\mathcal{I}_1 \land \forall i: \mathcal{I}_2(i)$),
it is clear that only one thread can be in the
critical section at the same time.
% For those who read through \autoref{ch:peterson}
% $\mathcal{I}_1 \land \forall i: \mathcal{I}_2(i)$ is an inductive invariant.

To see that invariant $\mathcal{I}_1$ is maintained, note that
\texttt{!}<{p}> = <{True}> upon entry of \texttt{test\_and\_set}
(because of the condition on the <{while}> loop that the
\texttt{test\_and\_set} method is invoked in).
There are two cases:
\begin{enumerate}
\item \texttt{!}<{s}> is <{False}> upon entry to \texttt{test\_and\_set}.
Then upon exit \texttt{!}<{p}> = <{False}> and \texttt{!}<{s}> = <{True}>,
maintaining the invariant.
\item \texttt{!}<{s}> is <{True}> upon entry to \texttt{test\_and\_set}.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $\mathcal{I}_1$ is also easy to verify for exiting the critical section
because we can assume, by the induction hypothesis, that $\textit{private}[i]$ is
<{True}> just before exiting the critical section.
Invariant $\mathcal{I}_2(i)$ is obvious as (i) thread <{i}> only proceeds to the critical
section if \textit{private}[<{i}>] is <{False}>, and (ii) no other thread modifies
\textit{private}[<{i}>].

@@

\autoref{fig:tas} shows a lock implementation based on a shared
variable and a private variable for each thread.   The private
variables themselves are actually implemented as shared variables,
but they are accessed only by their respective threads.
A thread usually does not keep explicit track of whether it has a lock
or not, because it is implied by the control flow of the program---a
thread implicitly \emph{knows} that when it is executing in a critical
section it has the lock.
There is no need to keep \textit{private} as a shared
variable---we only did so to be able to show and check the invariants.
\autoref{fig:tas2} shows a more straightforward implementation of a spinlock.
% The solution is similar to the na\"{i}ve solution of \autoref{fig:naivelock},
% but uses test-and-set to check and set the lock variable atomically.
The lock is also cleared in an atomic statement to prevent a data race.
This approach is general for any number of threads.

@@

\autoref{fig:ticket} similarly uses two variables for a lock,
\textit{counter} and \textit{dispenser}.  When a thread acquires the
lock, it fetches the current dispenser value and increments it
modulo \texttt{MAX\_THREADS}, all in one atomic operation.
In practice, \texttt{MAX\_THREADS} would
be a number like $2^{32}$ or $2^{64}$, but since the Harmony model checker
checks every possible state, limiting \texttt{MAX\_THREADS}
to a small number
significantly reduces the time to model check a Harmony program.
Plus it is easier to check that it fails when you run it with more than
\texttt{MAX\_THREADS} threads.
You can test the implementation using the command
\texttt{harmony -m synch=ticket code/cssynch.hny}.
To see it fail, try
\texttt{harmony -c NTHREADS=10 -m synch=ticket code/cssynch.hny}.

@@

\begin{figure}
\begin{code}
<{:input:../code/locksusp.hny}>
\end{code}
\caption{\harmonylink{modules/synchS.hny} The lock implementation in the \texttt{synchS} module uses suspension}
\label{fig:suspension}
\end{figure}

@@

\autoref{fig:suspension} shows the lock interface using suspension.
It is implemented as follows:
\begin{itemize}
\item A lock maintains both a boolean indicating whether the
lock is currently acquired and a list of contexts of threads that want to
acquire the lock.
\item
<{acquire}>()
\index{acquire}%
acquires the lock if available and suspends the invoking thread if not.
In the latter case, the context of the thread is added to the end of the list of contexts.
Note that \textbf{stop} is called within an atomic statement block---this is
the only exception to such an atomic statement block running to completion.
While the
thread is running no other threads can run, but when the thread suspends itself
other threads can run.
\item
<{release}>()
\index{release}%
checks to see if any threads are waiting to acquire the lock.
If so, it uses the <{head}> and <{tail}>
methods from the \texttt{list} module (see \autoref{ap:module})
to resume the first thread that got
suspended and to remove its context from the list.
\end{itemize}
Selecting the first thread is a design choice.  Another implementation could
have picked the last one, and yet another implementation could have used
<{choose}> to pick an arbitrary one.  Selecting the first is a common
choice in lock implementations as it prevents starvation.

@@

\section*{Exercises}
\begin{problems}
\item
Run \autoref{fig:incfixed} using (i) <{synch}> and (ii) \texttt{synchS}.
Report how many states were explored by Harmony for each module.
\item \label{ex:xy} \autoref{fig:xy} shows a Harmony program with two variables <{x}>
(initially 0) and <{y}> (initially 100) that can be accessed through methods
\texttt{setX} and \texttt{getXY}.  An application invariant is that \texttt{getXY}
should return a pair that sums to 100.  Add the necessary synchronization code.
\item \label{ex:trylock} Implement \texttt{tryAcquire}(\textit{b}) as an additional
interface for both the <{synch}> and \texttt{synchS} modules.
This interface is like <{acquire}>(\textit{b}) but never blocks.  It
returns <{True}> if the lock was available (and now acquired) or <{False}>
if the lock was already acquired.
Hint: you do not have to change the existing code.
\item People who use an ATM often first check their balance and then withdraw
a certain amount of money not exceeding their balance.  A negative balance
is not allowed.  \autoref{fig:atm} shows two operations on bank accounts:
one to check the balance and one to withdraw money.
Note that all operations on accounts are carefully protected by a lock
(i.e., there are no data races).
The \texttt{customer}
method models going to a particular ATM and withdrawing money not exceeding
the balance.
Running the code through Harmony reveals that there is a bug.
It is a common type of concurrency bug known as \emph{Time Of Check Time Of
Execution} (TOCTOE).
\index{Time Of Check Time Of Execution}
\index{TOCTOE}
In this case, by the time the withdraw operation is performed,
the balance can have changed.
Fix the code in \autoref{fig:atm}.  Note, you should leave the
customer code the same. You are only allowed to change the
\texttt{atm\_} methods, and you cannot use the \textbf{atomically}
keyword.
\end{problems}

@@

We will first implement the queue as a linked list.
The implementation in \autoref{fig:queueconc}
uses the <{alloc}> module for dynamic allocation
\index{dynamic allocation}%
of nodes in the list using \texttt{malloc}() and \texttt{free}().
\texttt{malloc}(<{v}>) returns a new memory location initialized to <{v}>,
which should be released with \texttt{free}() when it is no longer in use.
The queue maintains a <{head}> pointer to the first element in the list
and a <{tail}> pointer to the last element in the list.
The <{head}> pointer is <{None}> if and only if the queue is empty.
(<{None}> is a special address value that is not the address of any
memory location.)

@@

\index{hand-over-hand locking}%
Each node has a lock, a value, and <{next}>, a pointer to the next node
(which is <{None}> for the <{inf}> node).
The <{_find(lst, v)}> helper method first
finds and locks two consecutive nodes <{before}> and <{after}>
such that <{before->data.value < v <= after->data.value}>.
It does so by performing something called \emph{hand-over-hand locking}.
It first locks the first node, which is the <{-inf}> node.
Then, iteratively, it obtains a lock on the next node and release the
lock on the last one, and so on, similar to climbing a rope
hand-over-hand.
Using \texttt{\_find}, the <{insert}>, \texttt{remove}, and \texttt{contains}
methods are fairly straightforward to implement.

@@

As with critical sections---when testing a concurrent data structure---we
need a specification.
For example, \autoref{fig:queues}(a) shows a sequential specification of a
queue in Harmony.
First, we can check if the queue implementation
in \autoref{fig:queueconc} meets the sequential
queue specification in \autoref{fig:queues}(a).
To check if the queue implementation meets the specification,
we need to see if any sequence of queue operations in the
implementation matches a corresponding sequence in the
specification.
We say that the implementation and the specification
have the same \emph{behavior}.
%
\autoref{fig:qtestseq} presents a test program that does
exactly this, for sequences of up to <{NOPS}>
queue operations.
It maintains two queues:
\begin{itemize}
\item[] \textit{specq}: the queue of the specification;
\item[] \textit{implq}: the queue of the implementation.
\end{itemize}
For each operation, the code first chooses whether to
perform a <{get}> or <{put}> operation.
In the case of a <{put}> operation, the code also
chooses which value to append to the queue.
All operations are performed on both the queue implementation
and the queue specification.
In the case of <{get}>, the results of the operation
on both the implementation and specification are checked against one another.

@@

\autoref{fig:queuebroken} contains an attempt at a queue implementation
where the queue is implemented by a linked list, with the first node
being a \texttt{dummy} node to prevent data races.  Each node in the
list contains a lock.  The <{put}>() method walks the list until
it gets to the last node, each time acquiring the lock to access the
node's fields.  When <{put}>() gets to the last node in the list,
it appends a new one.  The <{get}>() method locks the first (dummy)
node, removes the second from the list and frees it.  The method returns
the value from the removed node.

@@

The <{alloc}> module maintains a shared array <{pool}> that
it uses for dynamic allocation.  Apparently T2 tried to access
<{pool}>[0], but it does not exist, meaning that either it was not
yet allocated, or it had been freed since it was allocated.
When we look at the top half of the figure, we see that in fact thread
T1 allocated <{pool}>[0] in turn 2, but T3 freed it in turn 4.
Looking back down, we see that T1 executed <{thread(1)}> and has
since terminated, while T3 is executing <{thread(3)}>.

>>>>

However, if the model checker does not find any bugs, we do not
know for sure that the algorithm works for lists of more than
five numbers or for lists that have values other than the numbers
1 through 5.  Still, we would expect that the likelihood that there
are bugs remaining in the sorting algorithm is small.
That said, it would be easy to write a program
that sorts all lists of up to five numbers correctly but fails to
do so for a list of 6 numbers.  (Hint: simply use an <{if}>
statement.)

@@

The <{assert}> statement checks that the output is
correct.  If the program is correct, Harmony reports the size of the
``state graph'' (13 states in this case).  If not, Harmony also
reports what went wrong, typically by displaying a summary of an execution in
which something went wrong.

@@

Concurrent programming, aka multithreaded programming, involves multiple
threads
\index{thread}%
running in parallel while sharing variables.
\autoref{fig:progs} shows two programs.  Program (a) is sequential.
It sets <{shared}> to <{True}>, asserts that
<{shared = True}> and finally sets <{shared}> to <{False}>.
If you run the program through Harmony, it will not find any problems
because there is only one execution possible and 1) in that execution
the assertion does not fail and 2) the execution terminates.
Program (b) is concurrent---it executes methods <{f()}> and
<{g()}> in parallel.
If method <{g()}> runs and completes before <{f()}>, then
the assertion in <{f()}> will fail when <{f()}> runs.
This problem is an example of non-determinism: methods <{f()}>
and <{g()}> can run in either order.
In one order, the assertion fails, while in the other it does not.
But since Harmony will find all possible executions, it will find
the problematic one.

@@

\begin{problems}
\item \autoref{fig:incenter} shows an attempt at trying to fix the code of
    \autoref{fig:inc}.  Run it through Harmony and see what happens.  Based on
    the error output, describe in English what is wrong with the code by describing,
    in broad steps, how running the program can get into a bad state.
\item What if we moved line~5 of \autoref{fig:incenter} to after the <{if}>
    statement (between lines~7 and~8)?  Do you think that would work?  Run it through
    Harmony and describe either why it works or why it does not work.
\end{problems}

@@

\glsadd{sequential consistency}%
In 1981, Gary L.~Peterson came up with a beautiful solution to the mutual exclusion
problem, now known as ``Peterson's Algorithm''~\cite{Peterson81}.
The algorithm is an amalgam of the (incorrect) algorithms in
\autoref{fig:naiveflags} and \autoref{fig:naiveturn}, and is presented
in \autoref{fig:peterson}.
(The first line specifies that the <{flags}> and <{turn}>
variables are assumed to satisfy \emph{sequential consistency}---it prevents
Harmony from complaining about data races involving these variables,
explained in \autoref{ch:spinlock}.)

@@

The HVM starts in an initial state in which there is only
one thread (<{__init__()}>) and its program counter is~0.
A \emph{trace}
\index{trace}%
is a sequence of steps starting from the initial state,
resulting in a sequence of states.
When making a step, there are two sources of non-determinism
\index{non-determinism}%
in Harmony.
% \footnote{Actually, there is a third type called
% an \emph{interrupt}, discussed in \autoref{ch:interrupts}.}
One is when
there is more than one thread that can make a step.  The other is
when a thread executes a <{choose}> operation and there is
more than one choice.
Because there is non-determinism, there are multiple possible traces.
We call a state \emph{reachable}
\index{reachable state}%
if it is either the initial state
or it can be reached from the initial state through a finite trace.
A state is final
when there are no threads left to make state changes.

@@

When running Harmony, this assertion will fail.
You can check the HTML output to see what happened.
Suppose thread 0 is at the critical section, $\mathit{flags}[0]$ = true,
$\mathit{turn} = 1$, and thread 1 just finished the step in line 7,
setting $\mathit{flags}[1]$ to true.  Then $C(0)$ is violated.
But it suggests a new property:
$\mathcal{G}(i) =
\mathtt{thread}(i)@\mathtt{cs} \Rightarrow \mathcal{C}(i) \lor \mathtt{thread}(1-i)@10$.
That is, if thread <{i}> is at the critical section, then
either $\mathcal{C}(i)$ holds or thread $1-i$ is about to execute line 10.
\autoref{fig:petersonproof} formalizes $\mathcal{G}(i)$ in Harmony.
The label <{gate}> refers to line 10, that is,
the step that sets <{turn}> to $1-i$.
You can run \autoref{fig:petersonproof} to determine
that $\mathcal{G}(i)$ is an invariant for $i = 0, 1$.
Moreover, if $\mathcal{F}(i)$ and $\mathcal{G}(i)$ both hold for $i = 0, 1$,
then mutual exclusion holds.  We can show this using proof by
contradiction.  Suppose mutual exclusion is violated and thus both threads are in
the critical section.  By $\mathcal{F}$ it must be the case that both
<{flags}> are true.  By $\mathcal{G}$ and the fact that neither thread
is at label <{gate}>, we know that both $C(0)$ and $C(1)$ must hold.
This then implies that $\mathit{turn} = 0 \land \mathit{turn} = 1$, providing
the desired contradiction.

@@

First consider the case that thread~0 is at label <{cs}>.  If thread~0
were to take a step, then in the next state thread~0 would be no longer
at that label and $\mathcal{G}(0)$ would hold trivially over the next state.
Therefore we only need to consider a step by thread~1.
%
From $\mathcal{G}$ we know that one of the following three cases must hold before
thread~1 takes a step:
\begin{enumerate}
\item <{flags[1] = False}>;
\item <{turn = 0}>;
\item thread~1 is at label <{gate}>.
\end{enumerate}

Let us consider each of these cases.
We have to show that if thread~1 takes a step, then one of those
cases must hold after the step.
In the first case, if thread~1 takes a step, there are two possibilities:
either $flags[1]$ will still be <{False}> (in which case the first case
continues to hold), or $flags[1]$ will be <{True}>
and thread~1 will be at label <{gate}> (in which case the third case
will hold).
We know that thread~1 never sets <{turn}> to 1, so
if the second case holds before the step, it will also hold after the step.
Finally, if thread~1 is at label <{gate}> before the step, then after
the step <{turn}> will equal 0, and therefore the second case will hold
after the step.

@@

A cool anecdote is the following.  When the author of Harmony had to teach
Peterson's Algorithm, he refreshed his memory by looking at the Wikipedia
page.  The page claimed that the following predicate is invariant:
if thread <{i}> is in the critical section, then $\mathcal{C}(i)$ (i.e.,
$\mathcal{G}$ without the disjunct that thread $1-i$ is at label <{gate}>).
We already saw that this is not an invariant.
(The author fixed the Wikipedia page with the help of Fred B.~Schneider.)

@@

Harmony is not an object-oriented language like Python is.  In Python,
you can pass a reference to an object to a method, and that method
can then update the object.  In Harmony, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this, as mentioned in \autoref{ch:harmonymachine},
each shared variable has an \emph{address}, itself a Harmony value.
\index{address}%
If <{x}> is a shared variable, then the expression <{?x}> is the address of <{x}>.
If a variable contains an address, we call that variable a \emph{pointer}.
\index{pointer}%
If <{p}> is a pointer to a shared variable, then the
expression \texttt{!}<{p}> is the value of the shared variable.
In particular, <{!?x = x}>.
This is similar to how C pointers work (\texttt{*\string&}<{x = x}>).

@@

The code in \autoref{fig:lockspec} also uses the Harmony <{when}>
statement.  A <{when}> statement waits until a time in which
condition holds (not necessarily the first time) and then executes the
statement block.  The ``<{await condition}>'' statement is
the same as ``<{when condition: pass}>''.
Combined with the <{atomically}> keyword, the entire statement
is executed atomically at a time that the condition holds.

@@

To support implementing locks,
many CPUs have an atomic ``test-and-set'' (TAS)
\index{test-and-set}%
\index{TAS}%
operation.
Method <{test_and_set}> in \autoref{fig:tas} shows its specification.
Here <{s}> points to a shared boolean variable and <{p}>
to a private boolean variable, belonging to some thread.
The operation copies the value of the shared variable to the
private variable (the ``test'')
and then sets the shared variable to <{True}> (``set'').

\autoref{fig:tas} goes on to implement mutual exclusion for
a set of <{N}> threads.
The approach is called \emph{spinlock},
\index{spinlock}%
because each thread is ``spinning'' (executing a tight loop) until
it can acquire the lock.
The program uses $\mathtt{N}+1$ boolean variables.
Variable <{shared}> is initialized to <{False}> while
<{private}[i]}> for each thread <{i}> is initialized to <{True}>.

An important invariant, $\mathcal{I}_1$, of the program is that at any time at most
one of these variables is <{False}>.
Another invariant, $\mathcal{I}_2(i)$, is that if thread <{i}> is
in the critical section, then <{private}[i] = False}>.
Between the two (i.e., $\mathcal{I}_1 \land \forall i: \mathcal{I}_2(i)$),
it is clear that only one thread can be in the
critical section at the same time.
% For those who read through \autoref{ch:peterson}
% $\mathcal{I}_1 \land \forall i: \mathcal{I}_2(i)$ is an inductive invariant.

To see that invariant $\mathcal{I}_1$ is maintained, note that
<{!p = True}> upon entry of <{test_and_set}>
(because of the condition on the <{while}> loop that the
<{test_and_set}> method is invoked in).
There are two cases:
\begin{enumerate}
\item <{!s}> is <{False}> upon entry to <{test_and_set}>.
Then upon exit <{!p = False}> and <{!s = True}>,
maintaining the invariant.
\item <{!s}> is <{True}> upon entry to <{test_and_set}>.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $\mathcal{I}_1$ is also easy to verify for exiting the critical section
because we can assume, by the induction hypothesis, that $<{private}>[i]$ is
<{True}> just before exiting the critical section.
Invariant $\mathcal{I}_2(i)$ is obvious as (i) thread <{i}> only proceeds to the critical
section if <{private}>[<{i}>] is <{False}>, and (ii) no other thread modifies
<{private}>[<{i}>].

@@

\autoref{fig:tas} shows a lock implementation based on a shared
variable and a private variable for each thread.   The private
variables themselves are actually implemented as shared variables,
but they are accessed only by their respective threads.
A thread usually does not keep explicit track of whether it has a lock
or not, because it is implied by the control flow of the program---a
thread implicitly \emph{knows} that when it is executing in a critical
section it has the lock.
There is no need to keep <{private}> as a shared
variable---we only did so to be able to show and check the invariants.
\autoref{fig:tas2} shows a more straightforward implementation of a spinlock.
% The solution is similar to the na\"{i}ve solution of \autoref{fig:naivelock},
% but uses test-and-set to check and set the lock variable atomically.
The lock is also cleared in an atomic statement to prevent a data race.
This approach is general for any number of threads.

@@

\autoref{fig:ticket} similarly uses two variables for a lock,
<{counter}> and <{dispenser}>.  When a thread acquires the
lock, it fetches the current dispenser value and increments it
modulo <{MAX_THREADS}>, all in one atomic operation.
In practice, <{MAX_THREADS}> would
be a number like $2^{32}$ or $2^{64}$, but since the Harmony model checker
checks every possible state, limiting <{MAX_THREADS}>
to a small number
significantly reduces the time to model check a Harmony program.
Plus it is easier to check that it fails when you run it with more than
<{MAX_THREADS}> threads.
You can test the implementation using the command
\texttt{harmony -m synch=ticket code/cssynch.hny}.
To see it fail, try
\texttt{harmony -c NTHREADS=10 -m synch=ticket code/cssynch.hny}.

@@

\begin{figure}
\begin{code}
<{:input:../code/locksusp.hny}>
\end{code}
\caption{\harmonylink{modules/synchS.hny} The lock implementation in the <{synchS}> module uses suspension}
\label{fig:suspension}
\end{figure}

@@

\autoref{fig:suspension} shows the lock interface using suspension.
It is implemented as follows:
\begin{itemize}
\item A lock maintains both a boolean indicating whether the
lock is currently acquired and a list of contexts of threads that want to
acquire the lock.
\item
<{acquire()}>
\index{acquire}%
acquires the lock if available and suspends the invoking thread if not.
In the latter case, the context of the thread is added to the end of the list of contexts.
Note that \textbf{stop} is called within an atomic statement block---this is
the only exception to such an atomic statement block running to completion.
While the
thread is running no other threads can run, but when the thread suspends itself
other threads can run.
\item
<{release()}>
\index{release}%
checks to see if any threads are waiting to acquire the lock.
If so, it uses the <{head}> and <{tail}>
methods from the \texttt{list} module (see \autoref{ap:module})
to resume the first thread that got
suspended and to remove its context from the list.
\end{itemize}
Selecting the first thread is a design choice.  Another implementation could
have picked the last one, and yet another implementation could have used
<{choose}> to pick an arbitrary one.  Selecting the first is a common
choice in lock implementations as it prevents starvation.

@@

\section*{Exercises}
\begin{problems}
\item
Run \autoref{fig:incfixed} using (i) <{synch}> and (ii) <{synchS}>.
Report how many states were explored by Harmony for each module.
\item \label{ex:xy} \autoref{fig:xy} shows a Harmony program with two variables <{x}>
(initially 0) and <{y}> (initially 100) that can be accessed through methods
\texttt{setX} and \texttt{getXY}.  An application invariant is that \texttt{getXY}
should return a pair that sums to 100.  Add the necessary synchronization code.
\item \label{ex:trylock} Implement <{tryAcquire(b)}> as an additional
interface for both the <{synch}> and <{synchS}> modules.
This interface is like <{acquire(b)}> but never blocks.  It
returns <{True}> if the lock was available (and now acquired) or <{False}>
if the lock was already acquired.
Hint: you do not have to change the existing code.
\item People who use an ATM often first check their balance and then withdraw
a certain amount of money not exceeding their balance.  A negative balance
is not allowed.  \autoref{fig:atm} shows two operations on bank accounts:
one to check the balance and one to withdraw money.
Note that all operations on accounts are carefully protected by a lock
(i.e., there are no data races).
The \texttt{customer}
method models going to a particular ATM and withdrawing money not exceeding
the balance.
Running the code through Harmony reveals that there is a bug.
It is a common type of concurrency bug known as \emph{Time Of Check Time Of
Execution} (TOCTOE).
\index{Time Of Check Time Of Execution}
\index{TOCTOE}
In this case, by the time the withdraw operation is performed,
the balance can have changed.
Fix the code in \autoref{fig:atm}.  Note, you should leave the
customer code the same. You are only allowed to change the
\texttt{atm\_} methods, and you cannot use the \textbf{atomically}
keyword.
\end{problems}

@@

We will first implement the queue as a linked list.
The implementation in \autoref{fig:queueconc}
uses the <{alloc}> module for dynamic allocation
\index{dynamic allocation}%
of nodes in the list using <{malloc()}> and <{free()}>.
<{malloc(v}>) returns a new memory location initialized to <{v}>,
which should be released with <{free()}> when it is no longer in use.
The queue maintains a <{head}> pointer to the first element in the list
and a <{tail}> pointer to the last element in the list.
The <{head}> pointer is <{None}> if and only if the queue is empty.
(<{None}> is a special address value that is not the address of any
memory location.)

@@

\index{hand-over-hand locking}%
Each node has a lock, a value, and <{next}>, a pointer to the next node
(which is <{None}> for the <{inf}> node).
The <{_find(lst, v)}> helper method first
finds and locks two consecutive nodes <{before}> and <{after}>
such that <{before->data.value < v <= after->data.value}>.
It does so by performing something called \emph{hand-over-hand locking}.
It first locks the first node, which is the <{-inf}> node.
Then, iteratively, it obtains a lock on the next node and release the
lock on the last one, and so on, similar to climbing a rope
hand-over-hand.
Using \texttt{\_find}, the <{insert}>, <{remove}>, and \texttt{contains}
methods are fairly straightforward to implement.

@@

As with critical sections---when testing a concurrent data structure---we
need a specification.
For example, \autoref{fig:queues}(a) shows a sequential specification of a
queue in Harmony.
First, we can check if the queue implementation
in \autoref{fig:queueconc} meets the sequential
queue specification in \autoref{fig:queues}(a).
To check if the queue implementation meets the specification,
we need to see if any sequence of queue operations in the
implementation matches a corresponding sequence in the
specification.
We say that the implementation and the specification
have the same \emph{behavior}.
%
\autoref{fig:qtestseq} presents a test program that does
exactly this, for sequences of up to <{NOPS}>
queue operations.
It maintains two queues:
\begin{itemize}
\item[] <{specq}>: the queue of the specification;
\item[] <{implq}>: the queue of the implementation.
\end{itemize}
For each operation, the code first chooses whether to
perform a <{get}> or <{put}> operation.
In the case of a <{put}> operation, the code also
chooses which value to append to the queue.
All operations are performed on both the queue implementation
and the queue specification.
In the case of <{get}>, the results of the operation
on both the implementation and specification are checked against one another.

@@

\autoref{fig:queuebroken} contains an attempt at a queue implementation
where the queue is implemented by a linked list, with the first node
being a \texttt{dummy} node to prevent data races.  Each node in the
list contains a lock.  The <{put()}> method walks the list until
it gets to the last node, each time acquiring the lock to access the
node's fields.  When <{put()}> gets to the last node in the list,
it appends a new one.  The <{get()}> method locks the first (dummy)
node, removes the second from the list and frees it.  The method returns
the value from the removed node.

@@

The <{alloc}> module maintains a shared array <{pool}> that
it uses for dynamic allocation.  Apparently T2 tried to access
<{pool[0]}>, but it does not exist, meaning that either it was not
yet allocated, or it had been freed since it was allocated.
When we look at the top half of the figure, we see that in fact thread
T1 allocated <{pool[0]}> in turn 2, but T3 freed it in turn 4.
Looking back down, we see that T1 executed <{thread(1)}> and has
since terminated, while T3 is executing <{thread(3)}>.

Looking further at the stack traces, we can see that
T3 was in the process of executing
<{release(?q.lock)}> within <{get(?q}>).
T1 is currently executing <{acquire(?alloc.pool[0].lock)}>
within <{put(?q, 2)}>, but 
<{alloc.pool[0]}> does not exist.
The corresponding line of Harmony code is
<{atomically when not !binsema}>
in line 25 of the \texttt{sync} module.
